---
title: "Lab1"
author: "Caden McQuillen"
date: '2023-05-09'
output: html_document
---

## Task 1
```{r}
train_data <- data.frame( c(seq(1:6)), c(0,2, 0, 0, -1, 1), c(3,0, 1, 1, 0, 1), c(0,0, 3, 2, 1, 1), c("Red", "Red", "Red", "Green", "Green", "Red") )
colnames(train_data) <- c("Obs", "X1", "X2", "X3", "Y")

test_point <- c(0,0,0)

distances <- vector(length = nrow(train_data))

for (i in 1:nrow(train_data)){
  
  current_coords <- train_data[i, c(2,3,4)]
  current_distance <- as.numeric(sqrt( (test_point[1] - current_coords[1])^2 + (test_point[2] - current_coords[2])^2 + (test_point[3] - current_coords[3])^2 ))
  distances[i] <- current_distance
}

names(distances) <- train_data$Obs
distances
```


## Task 2

Our prediction for k = 1 would be Green since we take the point with the smallest distance from our test point which in this case is obs 5 and then classify our test point as whatever Y that is. 
```{r}

k <- 1
distances_df <- data.frame(as.numeric(names(distances)), sort(distances))
colnames(distances_df) <- c("Obs", "Distance")

k1_prediction <- train_data[ distances_df[k,]$Obs,]$Y
k1_prediction

```


## Task 3

We take the 3 smallest distances from our test point and get their correpsonding Ys. Then we classify our test point as whatever Y has the highest occurance. 
```{r}

k <- 3
distances_df <- data.frame(as.numeric(names(distances)), sort(distances))
colnames(distances_df) <- c("Obs", "Distance")

k3_predictions <- train_data[ distances_df[seq(1:k),]$Obs,]$Y
max(k3_predictions)


```

## Task 4
```{r}
library(class)

knn_prediction <- knn(train_data[,c(2,3,4)], test_point, train_data$Y, k = 3)
knn_prediction
```


## Task 5
If the true decision boundary is highly non-linear, then we would expect a lower K to be the best value. This is because we need more flexibility to reduce bias. Howver if we pick too small of a K, then we might increase variance too much. 

## Task 6
```{r}
library(plotly)
plot_df <- data.frame(rbind(train_data[,c(2,3,4)], test_point), c(train_data$Y, "Test") )
colnames(plot_df)[4] <- "Y"

fig <- plot_ly(plot_df, x = ~X1, y = ~X2, z = ~X3, color = ~Y, colors = c('green', 'red', "grey"))
fig
```

